Gradual Soundness: Lessons from Static Python

### 2021-10-20 ecoop-2022/paper.tex
ecoop: clarify 'flip' research direction
ecoop: sec2 intro

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the compelling
\SP{} language design in another context and measure performance.
Second, we want to lend focus to theoretical research on gradual typing.
The restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
may be worthwhile assumptions for theory going forward.
On the flip side, there are issues in today's \SP{} that might
be addressed by future research.

\section{A Tour of Static Python}
\label{s:tour}
%% purpose = SP looks like Pyre, adds soundness to certain types

@@

\begin{itemize}
  \item Pyre as a recommended, optional pre-check
  \item SP type checker
  \item custom bytecode
\end{itemize}

@@

\subsection{Migration Issues}

@@

>>>>

\newcommand{\shorturl}[2]{\href{#1#2}{#2}}
\newcommand{\SP}{Static Python}
\newcommand{\code}[1]{\texttt{#1}}

@@

We have written this paper with two audiences in mind.
First, we want to encourage system-builders to reproduce the compelling
\SP{} language design in another context and measure performance.
Second, we want to lend focus to theoretical research on gradual typing.
The restrictions that \SP{} adopts (relative to the ideal promise of gradual typing)
may be worthwhile assumptions for theory going forward.
On the flip side, future research may find a way to lift restrictions
that \SP{} currently adopts.

\section{\SP{} and the Cinder JIT}
\label{s:tour}
%% purpose = SP looks like Pyre, adds soundness to certain types

\SP{} is a type checker and bytecode compiler.
It validates programs written in ordinary Python syntax
and generates type-specific bytecode for the Cinder JIT compiler.
Cinder is a fork of CPython 3.8 that introduces a method-based
JIT and several bytecode instructions.

Running \SP{} code is no different than standard Python.
The main Cinder executable generates bytecode and performs
optimizations on the fly, same as CPython.

There are a few restrictions on classes
and a few optional types that programmers may wish to use,
but otherwise Python programs can run via \SP{}.
%% TODO awk

\subsection{Example Program}

\begin{figure}
  %% TODO replace with a real program
%  \begin{minipage}[t]{0.5\columnwidth}
  \begin{verbatim}
    from __static__ import PyDict

    def f(x: PyDict[str, int]):
      return x["A"]

    f({"A": 1}) # ==> 1
  \end{verbatim}
%  \end{minipage}\begin{minipage}[t]{0.5\columnwidth}
%  \begin{verbatim}
%    from __static__ import PyDict
%
%    def f(x: PyDict[str, int]):
%      check_args(x, PyDict)
%      return cast(x["A"], int)
%
%    invoke_function(f, {"A": 1}) # ==> 1
%  \end{verbatim}
%  %  ....
%  %  CHECK_ARGS
%  %  ....
%  %  INVOKE_FUNCTION
%  \end{minipage}
%  \begin{verbatim}
%    from __static__ import CheckedDict
%
%    def f(x: CheckedDict[str, int]):
%      return x["A"]
%
%    f(CheckedDict({"A": 1}))
%  \end{verbatim}
  \caption{A first \SP{} program}
  \label{fig:sp-example}
\end{figure}

\Cref{fig:sp-example} (left) presents a first example program
that defines a function \code{f} and calls it from typed code.
The syntax is normal Python with type annotations.
Only the import statement makes special use of \SP{};
it imports a type \code{PyDict} that describes Python dictionaries.

Although this program is typed, a different untyped module is free
to import the function \code{f} and invoke it with any sort of
argument.
For this reason, the compiled version of \cref{fig:sp-example} must
include a few run-time casts:
\begin{itemize}
  \item
    The function \code{f} must check that its arguments match the \code{PyDict} domain annotation.
    Concretely, this check accepts any dictionary.
  \item
    Because the domain check does not validate the contents of an incoming dictionary,
    the dict access must check that its result matches the function codomain.
\end{itemize}
These checks ensure a kind of type soundness.

\SP{} also uses types to drive optimizations.
Because the call to \code{f} in \cref{fig:sp-example} appears in typed code,
there is no need to check that it sends type-correct inputs.
Thus, the compiled call uses a Cinder bytecode instruction to skip past the
domain check and immediately enter the function body.

The tension between checks that enforce soundness and optimizations allowed by soundness
is a central theme of \SP{}.

@@

\begin{enumerate}
  \item Pyre as an optional, but recommended pre-check. (Not necessary for the model.)
  \item SP type checker
  \item Custom bytecode, Cinder JIT
\end{enumerate}

@@

\subsection{Checked Types}

@@

\subsection{Restrictions}

No overloading.
At most one method per class and name.
Ensures that receiver type does not affect dispatch.

Overrides must be compatible, if done in static code.
What's the reason?

\subsection{Additional Options}

strict modules

decorator to force inlining, for 1-line functions
